Regional-scale Base-flow Index of Gauged and Ungauged Basins in Drylands

### 2024-11-26 index.qmd
Updating figures

![Map of Arizona indicating US Geological Survey (USGS) streamgages used in this study. 8-digit HUC subbasin boundaries and physiographic regions shown.](images/StudyArea.png){#fig-study-area}

@@

![Mean annual BFI for the period of study (1991-2020) from instrumented stream flow data.](images/BFI_Instrumented.png){#fig-instrumented-bfi}

@@

![Trends in BFI over full period of record for instrumented sites used in this study. Red upward (blue downward) arrows indicate an increasing (decreasing) trend at a significance level of 5%. Nonsignificant trends are shown by colors of circles.](images/BFITrends.png){#fig-instrumented-trend}

![](images/BFITrends_significant.png)

@@

![Predicted mean annual BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8.png){#fig-bfi-huc}

>>>>

![Map of Arizona indicating US Geological Survey (USGS) streamgages used in this study. 8-digit HUC subbasin boundaries and physiographic regions shown.](images/StudyArea_20241126.png){#fig-study-area}

@@

![Mean annual BFI for the period of study (1991-2020) from instrumented stream flow data.](images/BFI_Instrumented_20241126.png){#fig-instrumented-bfi}

@@

![Trends in BFI over full period of record for instrumented sites used in this study. Red upward (blue downward) arrows indicate an increasing (decreasing) trend at a significance level of 5%. White circles represent sites with no statistically significant trends.](images/BFITrends_20241126.png){#fig-instrumented-trend}

@@

![Predicted mean annual BFI values for 8-digit HUC (1991-2020)](images/BFI_HUC8_20241126.png){#fig-bfi-huc}

### 2024-11-26 index.tex
Updating figures

\includegraphics{images/StudyArea.png}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1892}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3784}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1081}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3243}}@{}}
\caption{\textbf{Table 1.} Basin-characteristic variables used as
predictors in XGBoost model.}\label{tbl-predictors}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Source
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Geoprocessing
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Source
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Geoprocessing
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Hydroclimate & Precipitation & PRISM & Basin average \\
& Mean Temperature & PRISM & Basin average \\
& Average Evapotranspiration & MODIS & Basin average \\
Geospatial & Elevation & DEM & Basin average \\
& Area & DEM & Basin average \\
& Slope & DEM & Basin average \\
& Aspect & DEM & Percent areal coverage \\
& Land Cover & NLCD & Percent areal coverage \\
& Hydrologic Soil Group & SSURGO & Percent areal coverage \\
& Geology & USGS & Percent areal coverage \\
& Karst & USGS & Percent areal coverage \\
\end{longtable}

@@

Statisical analyses were performed on mean annual BFI values of
instrumented streamgages. We examined the presence of temporal trends in
mean annual BFI using the Mann-Kendall nonparametric trend test
(Kendall, 1970; Mann, 1945). The Mann-Kendall was used to determine
trends in BFI and trend significance of gauged reaches during their
period of record.

@@

\includegraphics{images/BFI_Instrumented.png}

@@

The long-term average BFI for the 245 gauged reaches across Arizona is
illustrated in \textbf{?@fig-instrumented-bfi.The} highest BFI values
(\textgreater0.9) are found along the Grand Canyon in northwestern
Arizona. The highly karstic geology of this region facilitates the rapid
movement of subsurface flow to surface water and spring outlets
(Chambless et al., 2023). Relatively high BFI values (\textgreater0.8)
are found at the spring-fed headwaters of the Verde River (Del Rio
Spring) and the spring-fed headwaters of Fossil Creek. These results are
consistent with interpolated BFI values reported by Wolock (2003).

@@

\includegraphics{images/BFITrends.png}

@@

\caption{\label{fig-instrumented-trend}Trends in BFI over full period of
record for instrumented sites used in this study. Red upward (blue
downward) arrows indicate an increasing (decreasing) trend at a
significance level of 5\%. Nonsignificant trends are shown by colors of
circles.}

@@

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.3797}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0759}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0886}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1013}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0886}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0886}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0886}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0886}}@{}}
\caption{Table 2. Performance of model predictions for BFI for all sites
split by various classifications. n is number of sites,
R\textsuperscript{2} is the coefficient of determination of a linear
regression, MSE is mean-squared-error, RMSE is root-mean-squared-error,
MAE is mean-absolute-error, NSE is Nash-Sucliffe efficiency, and pbias
is percent bias.}\label{tbl-performance}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
n
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
R\textsuperscript{2}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
MSE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
RMSE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
MAE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
NSE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
pbias
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
n
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
R\textsuperscript{2}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
MSE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
RMSE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
MAE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
NSE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
pbias
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Climate - Monsoon Dominated & 3345 & 0.707 & 0.0166 & 0.129 & 0.077 &
0.694 & -12.6 \\
Climate - Snowmelt Dominated & 6152 & 0.723 & 0.017 & 0.131 & 0.094 &
0.715 & -3.7 \\
PhysRegion - Basin\&Range & 7754 & 0.749 & 0.018 & 0.133 & 0.089 & 0.741
& -5.9 \\
PhysRegion - CO Plateau & 1743 & 0.806 & 0.013 & 0.115 & 0.081 & 0.801 &
-4.4 \\
Climate - Warm-Wet & 1613 & 0.690 & 0.015 & 0.124 & 0.081 & 0.679 &
-9.9 \\
Climate - Warm-Dry & 3134 & 0.758 & 0.023 & 0.151 & 0.097 & 0.745 &
-8.1 \\
Climate - Cool-Wet & 3134 & 0.710 & 0.013 & 0.115 & 0.085 & 0.708 &
-2.5 \\
Climate - Cool-Dry & 1616 & 0.809 & 0.014 & 0.118 & 0.082 & 0.803 &
-5 \\
Slope - High & 4398 & 0.753 & 0.014 & 0.120 & 0.084 & 0.746 & -4.4 \\
Slope - Low & 5099 & 0.759 & 0.019 & 0.138 & 0.090 & 0.750 & -6.9 \\
\end{longtable}

@@

\includegraphics{images/BFI_HUC8.png}

>>>>

\includegraphics{images/StudyArea_20241126.png}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1892}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3784}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1081}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3243}}@{}}
\caption{Basin-characteristic variables used as predictors in XGBoost
model.}\label{tbl-predictors}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Source
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Geoprocessing
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Variable
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Source
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Geoprocessing
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Hydroclimate & Precipitation & PRISM & Basin average \\
& Mean Temperature & PRISM & Basin average \\
& Average Evapotranspiration & MODIS & Basin average \\
Geospatial & Elevation & DEM & Basin average \\
& Area & DEM & Basin average \\
& Slope & DEM & Basin average \\
& Aspect & DEM & Percent areal coverage \\
& Land Cover & NLCD & Percent areal coverage \\
& Hydrologic Soil Group & SSURGO & Percent areal coverage \\
& Geology & USGS & Percent areal coverage \\
& Karst & USGS & Percent areal coverage \\
\end{longtable}

@@

Statisical analyses were performed on mean annual BFI values of
instrumented streamgages. We examined the presence of temporal trends in
mean annual BFI using the Mann-Kendall nonparametric trend test
(Kendall, 1970; Mann, 1945). The Mann-Kendall test was used to determine
trends in BFI and trend significance of gauged reaches.

@@

\includegraphics{images/BFI_Instrumented_20241126.png}

@@

The long-term average BFI for the 244 gauged reaches across Arizona is
illustrated in Figure~\ref{fig-instrumented-bfi} . The highest BFI
values (\textgreater0.9) are found along the Grand Canyon in
northwestern Arizona. The highly karstic geology of this region
facilitates the rapid movement of subsurface flow to surface water and
spring outlets (Chambless et al., 2023). Relatively high BFI values
(\textgreater0.8) are found at the spring-fed headwaters of the Verde
River (Del Rio Spring) and the spring-fed headwaters of Fossil Creek.
These results are consistent with interpolated BFI values reported by
Wolock (2003).

@@

\includegraphics{images/BFITrends_20241126.png}

@@

\caption{\label{fig-instrumented-trend}Trends in BFI over full period of
record for instrumented sites used in this study. Red upward (blue
downward) arrows indicate an increasing (decreasing) trend at a
significance level of 5\%. White circles represent sites with no
statistically significant trends.}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.4348}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0725}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1014}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1014}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1449}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1449}}@{}}
\caption{Comparison of trends for BFI for all sites split by various
classifications. Only sites with a significant (\(\rho \le 0.05\)) trend
are included here as established by a Mann-Kendall test for monotonic
trends across the full period of record. n is the number of sites,
n\_pos (n\_neg) is the number of sites with positive (negative) trends,
perc\_pos (perc\_neg) is the percentage of n with a positive (negative)
trend.}\label{tbl-trends}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Classification Group
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
n
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
n\_pos
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
n\_neg
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
perc\_pos
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
perc\_neg
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Classification Group
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
n
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
n\_pos
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
n\_neg
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
perc\_pos
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
perc\_neg
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Climate - Monsoon Dominated & 94 & 10 & 23 & 0.106 & 0.245 \\
Climate - Snowmelt Dominated & 150 & 19 & 17 & 0.127 & 0.113 \\
PhysRegion - Basin\&Range & 190 & 26 & 33 & 0.137 & 0.174 \\
PhysRegion - CO Plateau & 54 & 3 & 7 & 0.056 & 0.130 \\
Climate - Warm-Wet & 31 & 3 & 6 & 0.097 & 0.194 \\
Climate - Warm-Dry & 81 & 15 & 16 & 0.185 & 0.198 \\
Climate - Cool-Wet & 83 & 6 & 11 & 0.072 & 0.133 \\
Climate - Cool-Dry & 49 & 5 & 7 & 0.102 & 0.143 \\
Slope - High & 111 & 12 & 15 & 0.108 & 0.135 \\
Slope - Low & 133 & 17 & 25 & 0.128 & 0.188 \\
\end{longtable}

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.3797}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0759}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0886}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1013}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0886}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0886}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0886}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.0886}}@{}}
\caption{Performance of model predictions for BFI for all sites split by
various classifications. n is number of observatios,
R\textsuperscript{2} is the coefficient of determination of a linear
regression, MSE is mean-squared-error, RMSE is root-mean-squared-error,
MAE is mean-absolute-error, NSE is Nash-Sucliffe efficiency, and pbias
is percent bias.}\label{tbl-performance}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
n
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
R\textsuperscript{2}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
MSE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
RMSE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
MAE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
NSE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
pbias
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
n
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
R\textsuperscript{2}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
MSE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
RMSE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
MAE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
NSE
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
pbias
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Climate - Monsoon Dominated & 3345 & 0.707 & 0.0166 & 0.129 & 0.077 &
0.694 & -12.6 \\
Climate - Snowmelt Dominated & 6152 & 0.723 & 0.017 & 0.131 & 0.094 &
0.715 & -3.7 \\
PhysRegion - Basin\&Range & 7754 & 0.749 & 0.018 & 0.133 & 0.089 & 0.741
& -5.9 \\
PhysRegion - CO Plateau & 1743 & 0.806 & 0.013 & 0.115 & 0.081 & 0.801 &
-4.4 \\
Climate - Warm-Wet & 1613 & 0.690 & 0.015 & 0.124 & 0.081 & 0.679 &
-9.9 \\
Climate - Warm-Dry & 3134 & 0.758 & 0.023 & 0.151 & 0.097 & 0.745 &
-8.1 \\
Climate - Cool-Wet & 3134 & 0.710 & 0.013 & 0.115 & 0.085 & 0.708 &
-2.5 \\
Climate - Cool-Dry & 1616 & 0.809 & 0.014 & 0.118 & 0.082 & 0.803 &
-5 \\
Slope - High & 4398 & 0.753 & 0.014 & 0.120 & 0.084 & 0.746 & -4.4 \\
Slope - Low & 5099 & 0.759 & 0.019 & 0.138 & 0.090 & 0.750 & -6.9 \\
\end{longtable}

Consistently negative pbias values across all classifications suggest an
under-prediction of BFI.

@@

\includegraphics{images/BFI_HUC8_20241126.png}
